---
title: "Homework 2"
author: "Reka Solymosi"
date: "26 September 2016"
output: html_document
---

##Question 1: Read in data
There is lots of very interesting data online. One example is data on football banning orders, which is collected by the Home Office, and is available [here](https://data.gov.uk/dataset/football-banning-orders). As most data, it's a bit messy, so I've done some of the cleaning for you. So you can download a .csv of some clean data that you should use for this part of the homework from Blackboard, avilable [here](). 

Download this data set, and read this dataset into R.

Enter the code you used for this below (don't forget to annotate): 

```{r, eval=FALSE}

#enter your code here




```
*(Hint: remember to give your data object a memorable name)*

##Question 2: Describe the dataset

**How many variables and cases does this dataset have?** 
Paste the code used to get the answers below: 

```{r, eval=FALSE}

#enter your code here




```
And the answer is: 
*(Hint: check its DIMensions)*

**What are the variables in this dataset? What class are they?**

Paste the code used to get the answers below: 

```{r, eval=FALSE}

#enter your code here




```
And the answer is: 
*(Hint: you can apply a function of checking class to the whole dataframe)*


**What are the names of the variables?** 
Paste the code used to get the answers below: 

```{r, eval=FALSE}

#enter your code here




```
And the answer is: 
*(Hint: you want the names())*

## Question 3: Mean

**What is the average (mean) number of banning orders accross supporters of all clubs in the dataset?**

Paste the code used to get the answers below: 

```{r, eval=FALSE}

#enter your code here




```

And the answer is: 

##Question 4: Filter 

Use filtering to answer the following questions: 

**What is the average (mean) number of banning orders accross only supporters of Premier league clubs?**

Paste the code used to get the answers below: 

```{r, eval=FALSE}

#enter your code here




```

And the answer is:

**What is the average (mean) number of banning orders accross only supporters of League Two clubs?**

Paste the code used to get the answers below: 

```{r, eval=FALSE}

#enter your code here




```

And the answer is:

##Question 5: Discussion

**Produce a simple histogram of the Banning.Orders variable. **

*(Hint: HISTogram)*

Any discernible patterns? Discuss.

Any code goes here: 
```{r, eval=FALSE}

#enter your code here




```

**Produce summary statistics of the Banning.Orders variable by League.of.the.Club.Supported. **

*(Hint: you want to DESCRIBE banning orders BY leage)*

Any discernible patterns? Discuss.

Any code goes here: 
```{r, eval=FALSE}

#enter your code here




```

