---
title: "Homework template"
author: "Reka Solymosi"
date: "23 September 2016"
output: html_document
---

##Question 1: 
<!-- What was the question you asked? -->

how many burglaries were there?

###Why are we asking this question:
<!-- Justify why you would want to ask this question, what could the answer be useful for -->

i want to know
###How we want to answer this question:
<!-- Here describe the approach you will take to answer this question -->

frequency table of crime types look at number 

###This is our code

```{r, eval=FALSE}
#paste your R code here


#end of code here
```

###This is our answer
<!-- So what's your answer? -->

Type here

##Question 2: 
<!-- What was the question you asked? -->

Type here

###Why are we asking this question:
<!-- Justify why you would want to ask this question, what could the answer be useful for -->

Type here

###How we want to answer this question:
<!-- Here describe the approach you will take to answer this question -->

Type here

###This is our code

```{r, eval=FALSE}
#paste your R code here



#end of code here
```

###This is our answer
<!-- So what's your answer? -->

Type here

##Question 3: 
<!-- What was the question you asked? -->

Type here

###Why are we asking this question:
<!-- Justify why you would want to ask this question, what could the answer be useful for -->

Type here

###How we want to answer this question:
<!-- Here describe the approach you will take to answer this question -->

Type here

###This is our code

```{r, eval=FALSE}
#paste your R code here



#end of code here
```

###This is our answer
<!-- So what's your answer? -->

Type here

